


```{r}
phe_num <- inputs$phe_num[i]
cross_num <- inputs$cross_num[i]
res_list <- list()
for (i in 1:22){
  res_list[[i]] <- readr::read_table(paste0("~/research/ukb-intervals/results/pheno", phe_num, "/summary_ukb_cross", cross_num, "_variance_chr", i, ".txt"), col_names = FALSE, skip = 2)
}
# fix column names
for (chr in 1:22){
  colnames(res_list[[chr]]) <- paste0("chr", chr, "_", colnames(res_list[[chr]]))
}

alpha = 0.1
# calc sd
dat <- do.call("cbind", res_list) %>% 
  tibble::as_tibble() %>%
  dplyr::mutate(total_var = rowSums(.)) %>%
  dplyr::mutate(total_sd = sqrt(total_var)) %>%
  dplyr::select(total_sd) %>%
  dplyr::mutate(interval_half_width = qnorm(alpha / 2, lower.tail = FALSE) * total_sd)
```

# Reading plink allele score outputs

We now need to read the allele score outputs for the test set.

To ensure that we're reading the files accurately, we'll also read the fam file, which contains the trait values.

```{r}
#read test indicator file
test_indicator <- readr::read_table(file = paste0("~/research/ukb-intervals/test_indicator_files/test_indicator_pheno_", phe_num, "_cross_", cross_num, "_ntest1000.txt"), col_names = FALSE)
```

```{r}
# read fam file
fam <- readr::read_table(file = "~/research/predictionProject/plink_file/ukb/chr1.fam", col_names = FALSE) %>% 
  dplyr::mutate(test_indic = test_indicator$X1)
## define col_num
col_num <- phe_num * 5 + (cross_num %% 5) + 1 
## 
fam2 <- fam %>%
  dplyr::select(1, 2, dplyr::all_of(col_num), 131) %>%
  dplyr::filter(test_indic == 1)
```

```{r}
# read the profile files outputted by plink score allele
scores <- list()
for (i in 1:22){
  fn <- paste0("~/research/ukb-intervals/results/pheno", phe_num, "/predicted_ukb_cross", cross_num, "_chr", i, "_auto.dbslmm.txt.profile")
  scores[[i]] <- readr::read_table(fn, col_names = TRUE)
  colnames(scores[[i]]) <- paste0(colnames(scores[[i]]), "_chr", i)
}
scores2 <- lapply(X = scores, FUN = function(x){
  x %>% dplyr::select(6) # choose the SCORESUM column
})
scores_df <- do.call("cbind", scores2) %>% tibble::as_tibble() %>%
  dplyr::mutate(sum_scores = rowSums(.))
s2 <- scores_df[1:1000,] %>%
  dplyr::select(sum_scores)
fam3 <- fam2 %>%
  dplyr::mutate(predicted_value = s2$sum_scores) %>%
  dplyr::mutate(interval_half_width = dat$interval_half_width)
```

```{r}
fam3 %>%
  dplyr::rename(true_pheno = 3) %>%
  dplyr::mutate(diff = abs(true_pheno - predicted_value),
                in_interval = interval_half_width >= diff) %>%
  dplyr::summarise(coverage = mean(in_interval))
```

## Use the validation set

We borrow code from Sheng Yang's file, "comparisonProject/code/03_internal/06_summ_ukb_c.R".


```{r}
library(plyr)
library(bigreadr)
library(Metrics)
library(optparse)



##############################
### summarize the result of validate
##############################
## summ function
summPheno <- function(test_pheno, chr_path, lm_coef, type){
  
  require(bigreadr)
  require(Metrics)
  pred_tot <- vector()
  for(chr in 1: 22){
    pred_chr_str <- paste0(chr_path, chr, type, ".profile.gz")
    if (file.exists(pred_chr_str)){
      pred_chr <- fread2(pred_chr_str, header = T)[, 6]
      pred_tot <- cbind(pred_tot, pred_chr)
    } else {
      cat (paste0("chr:", chr, " fail!\n"))
    }
  }
  pheno_tot <- rowSums(pred_tot) 
  pred <- cbind(1, pheno_tot) %*% lm_coef
  r2 <- cor(test_pheno, pred[, 1])^2
  mse <- mse(test_pheno, pred[, 1])
  result <- list(r2, mse, pred)
  return(result)
}

## val function
summPheno_val <- function(valid_pheno, chr_path, type){
  require(bigreadr)
  require(Metrics)

  pred_tot <- vector()
  for(chr in 1: 22){
    pred_chr_str <- as.character(paste0(chr_path, chr, type, ".profile.gz"))
    
    if (file.exists(pred_chr_str)){
      pred_chr <- read.table(pred_chr_str, header = T)[, 6]
      pred_tot <- cbind(pred_tot, pred_chr)
    } else {
      cat (paste0("chr:", chr, " fail!\n"))
    }
  }
  pheno_tot <- rowSums(pred_tot)
  lm_coef <- coef(lm(valid_pheno~pheno_tot))
  return(lm_coef)
}

comp_str <- "/net/mulan/disk2/yasheng/comparisonProject/"
method <- "DBSLMM"
situ <- "_ukb"
situ2 <- ""
m <- 1
pheno_path <- paste0(comp_str, "05_internal_c/")
val_pheno <- fread2(paste0(comp_str, "03_subsample/continuous/pheno", 
                           phe_num, "/val/ukb/02_pheno_c.txt"))[, 1]
# c(1: length(method))

    test_pheno <- fread2(paste0(comp_str, "02_pheno/03_test_c/pheno_pheno", 
                                phe_num, ".txt"), select = cross_num)[, 1]
    test_pheno_na <- test_pheno[!is.na(test_pheno)]
    
    #val_chr_path <- paste0(pheno_path, "pheno", pheno[p], "/", method[m], 
    #                       "/val", situ[m], "_cross", cross, "_chr") 
    val_chr_path <- paste0(pheno_path, "pheno", phe_num, "/", "DBSLMM", 
                          "/val", "_ukb", "_cross", cross_num, "_chr")
    chr_path <- paste0(pheno_path, "pheno", phe_num, "/", "DBSLMM", 
                       "/pred", situ[1], "_cross", cross_num, "_chr") 
    
    lm_coef <- try(summPheno_val(val_pheno, val_chr_path, situ2[m]), silent = T)
    temp_res <- try(summPheno(test_pheno_na, chr_path, lm_coef, situ2[m]), silent = T)

```

```{r}
fam3 %>%
  dplyr::rename(true_pheno = 3) %>%
  dplyr::mutate(validated_predicted_value = temp_res[[3]][1:1000, 1]) %>%
  dplyr::mutate(diff = abs(true_pheno - validated_predicted_value),
                in_interval = interval_half_width >= diff) %>%
  dplyr::summarise(coverage = mean(in_interval))

```


## Using a calibration set


We'll choose the first 500 subjects of the "test" set to be a calibration set. The last 500 will be the "true test" set.

```{r}
calibration_set <- fam3[1:500, ] %>%
  dplyr::rename(true_pheno = 3) %>%
  dplyr::mutate(validated_predicted_value = temp_res[[3]][1:500, 1]) %>%
  dplyr::mutate(diff = abs(true_pheno - validated_predicted_value),
                in_interval = interval_half_width >= diff) %>%
  dplyr::mutate(multiplicative_factor = diff / interval_half_width) %>%
  dplyr::arrange(multiplicative_factor) %>% 
  dplyr::mutate(overall_multiplicative_factor = .$multiplicative_factor[450]) %>%
  dplyr::mutate(adjusted_interval_half_width = interval_half_width * overall_multiplicative_factor) %>%
  dplyr::mutate(in_adjusted_interval = adjusted_interval_half_width >= diff)
calibration_set %>% dplyr::summarise(coverage = mean(in_adjusted_interval))

```

```{r}
fam3[501:1000, ] %>%
  dplyr::rename(true_pheno = 3) %>%
  dplyr::mutate(validated_predicted_value = temp_res[[3]][501:1000, 1]) %>%
  dplyr::mutate(diff = abs(true_pheno - validated_predicted_value)) %>%
  dplyr::mutate(adjusted_interval_half_width = interval_half_width * calibration_set$overall_multiplicative_factor[1]) %>%
  dplyr::mutate(in_adjusted_interval = adjusted_interval_half_width >= diff) %>%
  dplyr::summarise(coverage = mean(in_adjusted_interval))

```

