---
title: "Creating file for subsetting to training and test sets"
author: "Frederick Boehm"
date: "1/6/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Our goal here is to produce a plain text file with one column of integers. Those 
integers correspond to the row numbers (minus one, for C++) of the subjects that get 
assigned to the training set.

To accomplish this goal, we need the
number of subjects. Once we have that number, we can specify the proportion to be in 
the training set and use R's pseudorandom sampling to partition subjects.


```{r}
# which column in the fam file?
fam <- readr::read_delim("~/research/ukb-intervals/dat/plink_files/ukb/chr1.fam", col_names = FALSE)
fam_nona <- fam[!is.na(fam$X6), ] # change column name for other traits
n_total <- nrow(fam_nona) 
# Does the fam file have column names???
# No! So every line corresponds to a subject!
```

Now that we have `n_total`, the total number of subjects in the fam file, after discarding those with 
missing trait values for teh trait of interest, we 
proceed by specifying the proportion of the subjects that will be in the training set.

```{r}
training_proportion <- 0.8
set.seed(2022-01-06)
cpp_indices <- sample(0:(n_total - 1), size = floor(training_proportion * n_total))
r_indices <- cpp_indices + 1
readr::write_delim(x = tibble::as_tibble(cpp_indices), 
                   quote = "none",
                   file = "training_indices.txt", # this is not hardcoded into dbslmm.cpp
                   col_names = FALSE)
readr::write_delim(x = tibble::as_tibble(fam_nona$X1[1 + cpp_indices]), 
                   file = "training_ids.txt", 
                   quote = "none",
                   col_names = FALSE)
```

```{r}
# determine test indices
cpp_indices_test <- setdiff(0:(n_total - 1), cpp_indices)
readr::write_delim(x = tibble::as_tibble(cpp_indices_test), 
                   quote = "none",
                   file = "test_indices.txt", # this is not hardcoded into dbslmm.cpp
                   col_names = FALSE)

```

```{r}
# write the training fam file for this trait
fam_nona[!(1:nrow(fam_nona) %in% r_indices), 6:ncol(fam_nona)] <- NA
readr::write_delim(x = fam_nona, 
                   file = "~/research/ukb-intervals/dat/plink_files/ukb-gemma-training/pheno1/chr1.fam", # change path for different traits
                   quote = "none",
                   col_names = FALSE)

```
