---
title: "Make index files for test set"
author: "Frederick J. Boehm"
date: "4/6/2022"
output: html_document
---

The goal is to create, for every cross-pheno combination, a file containing 
the indices of subjects that belong to the test set. We'll use the phenotype files 
that Sheng Yang created for his study of UKB traits.

```{r}
# read the fam file from predictionProject dir.
library(magrittr)
fam <- readr::read_delim("/net/mulan/home/fredboe/research/predictionProject/plink_file/ukb/chr22.fam", col_names = FALSE)$X1 %>% tibble::as_tibble()
```

We have `fam`, a tibble with 337129 rows and 1 column. The one column contains the 
family ids for the subjects. These are not numbered consecutively, since they 
are taken from the larger data set (ie, with ~500k subjects).



```{r}
for (phe_num in 1:25){
for (cross_num in 1:5){  
    
    test_ids_filename <- paste0("/net/mulan/disk2/yasheng/comparisonProject/02_pheno/01_test_idx_c/idx_pheno", phe_num, "_cross", cross_num, ".txt")
    test_ids <- readr::read_table(test_ids_filename, col_names = FALSE)
    my_binary_vec <- (fam$value %in% test_ids$X1) %>% as.numeric()
    sum(my_binary_vec)
    length(my_binary_vec)
    ##
    fam %>%
      dplyr::mutate(in_test_set = my_binary_vec) %>%
      dplyr::filter(in_test_set == 1)
    test_ids
    outfn <- paste0("~/research/ukb-intervals/test_indicator_files/test_indicator_pheno_", phe_num, "_cross_", cross_num, ".txt")
    #readr::write_delim(x = my_binary_vec %>% tibble::as_tibble(), file = outfn, col_names = FALSE)
    # n_test defines the number of test subjects
    n_test <- 1000
    mbv <- numeric(length = length(my_binary_vec))
    mbv[which(my_binary_vec == 1)[1:n_test]] <- 1
    outfn <- paste0("~/research/ukb-intervals/test_indicator_files/test_indicator_pheno_", phe_num, "_cross_", cross_num, "_ntest", n_test, ".txt")
    readr::write_delim(x = mbv %>% tibble::as_tibble(), file = outfn, col_names = FALSE)
}
}
```

