---
title: "Quantile normalization in R"
author: "Fred Boehm"
date: "11/2/2021"
output: html_document
---

## Generate data

```{r}
dat <- rnorm(1000)
dat[901:1000] <- 0
```

## Xiang's method with `qqnorm` and `sample`

```{r}
xz_method <- qqnorm(dat, plot.it = FALSE)$x
```


## `ryouready::qqnorm_spss`

The function `ryouready::qqnorm_spss` has an argument, `ties.method`, for dealing with ties, while
`base::qqnorm` has no comparable argument.

It seems that we want to set `ties.method = "random"` for our analyses.

```{r}
ry_average <- ryouready::qqnorm_spss(dat, ties.method = "average")$y
```

```{r}
ry_random <- ryouready::qqnorm_spss(dat, ties.method = "random")$y
```

```{r}
tibble::tibble(xz_method = xz_method, ry_average = ry_average, ry_random = ry_random)
```

## Source code

It's easiest to see the differences when you compare the R code that defines the two functions, `stats:::qqnorm.default` and `ryouready::qqnorm_spss`. 

```{r}
stats:::qqnorm.default
```

```{r}
ryouready::qqnorm_spss
```

## Colophon


```{r}
if ( git2r::in_repository() ) {
        summary(git2r::last_commit()) 
 } else { 
        message("We are not in a git repository") 
 }
```

```{r}
devtools::session_info()
```
