---
title: "Cross-validation-plus for UKB traits"
author: "Frederick J. Boehm"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    number_sections: false
params:
  trait_type: "binary"
  nfold: 20
---

This Rmd is to be run after cv-plus.Rmd, as this file uses files produced by cv-plus.Rmd.

Our goal in this file is to look across traits, but within the six groups defined by `trait_type` and `nfold` values.



```{r}
trait_type <- params$trait_type
nfold <- params$nfold

library(magrittr)
```

## Load results files

```{r}
fbstr <- paste0("~/research/ukb-intervals/study_nfolds/", nfold, "-fold/")
res <- list()


for (trait_num in 1:25){# iterate over the 25 traits per group
    if (trait_type == "continuous") {
        output_dir <- paste0(fbstr, "05_internal_c/pheno", trait_num, "/cvplus/")
        res_file <- paste0(output_dir, "res.rds")
    } else {
        output_dir <- paste0(fbstr, "06_internal_b/pheno", trait_num, "/cvplus/")
        res_file <- paste0(output_dir, "res.rds")
    }
    if (!file.exists(res_file)) {
        stop("res_file doesn't exist")
    } else {
        res[[trait_num]] <- readRDS(res_file)[[1]] # need [[1]] since res file objects are stored as lists themselves
    }
}
```



## Boxplots of interval width

```{r, eval = TRUE}
for (i in seq_along(res)){
  res[[i]] <- res[[i]] %>%
    dplyr::mutate(trait_num = i)
}
dplyr::bind_rows(res) %>%
  ggplot2::ggplot() + 
  ggplot2::geom_boxplot(ggplot2::aes(x = trait_num, y = interval_width, group = trait_num))
```

## Incorporate estimated heritability of traits

```{r, eval = TRUE}
if (trait_type == "continuous"){
  h_tib <- readr::read_csv(paste0(fbstr, "05_internal_c/pheno25/herit/h2.csv"), 
                          col_names = FALSE) %>% 
            dplyr::select(-26) # drop 26, which is due to presence of trailing commas
} else {
  h_tib <- readr::read_csv(paste0(fbstr, "06_internal_b/pheno25/herit/h2.csv"), 
                          col_names = FALSE) %>%
            dplyr::select(- tidyselect::last_col()) # drop 26, which is due to presence of trailing commas
}

pp <- h_tib %>% 
  dplyr::summarise(dplyr::across(.fns = mean)) %>%
  tidyr::pivot_longer(seq_along(1:25)) %>%
  dplyr::rename(mean_herit = value) %>%
  dplyr::mutate(trait_num = 1:25) %>%
  dplyr::select(-name) %>%
  dplyr::select(trait_num, mean_herit) %>%
  dplyr::left_join(dplyr::bind_rows(res), by = "trait_num") 
pp2 <- pp %>%
  ggplot2::ggplot() + ggplot2::geom_boxplot(ggplot2::aes(x = mean_herit, y = interval_width, group = trait_num, colour = as.factor(trait_num), label = trait_num), show.legend = FALSE)
pp2 %>% plotly::ggplotly(tooltip = "label")
```


# Plot coverage vs (mean) heritability 

```{r, eval = TRUE}
ppsum <- pp %>%
  dplyr::group_by(trait_num) %>%
  dplyr::summarise(mean_herit = mean(mean_herit), coverage = mean(in_interval, na.rm = TRUE))
pp2 <- ppsum %>%
  ggplot2::ggplot() + ggplot2::geom_point(ggplot2::aes(x = mean_herit, y = coverage, colour = as.factor(trait_num))) 
pp2 %>%
  plotly::ggplotly()
```

## Check correlation between (mean) heritability and coverage


```{r, eval = TRUE}
cor(ppsum$mean_herit, ppsum$coverage)
```


# References





