---
title: "Cross-validation-plus for UKB traits"
author: "Frederick J. Boehm"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    number_sections: false
bibliography: ../Rmd/my.bib
params:
  trait_type: "continuous"
---

```{r}
library(magrittr)
```


## Custom R code

It should be straightforward to use DBSLMM outputs and fam files to write custom code for CV+.

Inputs needed: 

- fam file to get true values for $Y$. We need trait values for both the verification and training sets.    
- DBSLMM-based predicted values for a trait, for all subjects and for each of the five folds that Sheng created. 




```{r, echo = TRUE, message = FALSE}
(trait_type <- params$trait_type)
traits <- 1 # trait numbers to analyze
alpha <- 0.1
out <- list()
if (trait_type == "continuous") {
  res_file <- "res-continuous.rds"
} else {
  res_file <- "res-binary.rds"
}
if (!file.exists(res_file)) {
  res <- lapply(traits, function(x) {
    knitr::knit_child(
      'cv-plus-child.Rmd', 
      envir = environment(), 
      quiet = TRUE
    )
    get("out")
  })
  saveRDS(res, file = res_file)
} else {
  res <- readRDS(res_file)
}
```

```{r}
coverages <- lapply(X = res, 
                    FUN = function(mytib){
  mytib %>%
    dplyr::summarise(coverage = mean(in_interval))
       }
)
coverage_tib <- do.call("rbind", coverages) %>%
  tibble::as_tibble() %>%
  dplyr::mutate(trait_num = 1:length(res))
coverage_tib %>%
  gt::gt()

```

## Boxplots of interval width

```{r}
for (i in 1:length(res)){
  res[[i]] <- res[[i]] %>%
    dplyr::mutate(trait_num = i)
}
dplyr::bind_rows(res) %>%
  ggplot2::ggplot() + 
  ggplot2::geom_boxplot(ggplot2::aes(x = trait_num, y = interval_width, group = trait_num))
```

## Incorporate estimated heritability of traits

```{r, eval = FALSE}
if (trait_type == "continuous"){
  h_tib <- readr::read_csv("~/research/ukb-intervals/shell_scripts/h2.csv", 
                          col_names = FALSE) %>% 
            dplyr::select(-26) # drop 26, which is due to presence of trailing commas
} else {
  h_tib <- readr::read_csv("~/research/ukb-intervals/shell_scripts/h2-binary.csv", 
                          col_names = FALSE) %>%
            dplyr::select(- tidyselect::last_col()) # drop 26, which is due to presence of trailing commas
}

pp <- h_tib %>% 
  dplyr::summarise(dplyr::across(.fns = mean)) %>%
  tidyr::pivot_longer(1:length(traits)) %>%
  dplyr::rename(mean_herit = value) %>%
  dplyr::mutate(trait_num = traits) %>%
  dplyr::select(-name) %>%
  dplyr::select(trait_num, mean_herit) %>%
  dplyr::left_join(dplyr::bind_rows(res), by = "trait_num") 
pp2 <- pp %>%
  ggplot2::ggplot() + ggplot2::geom_boxplot(ggplot2::aes(x = mean_herit, y = interval_width, group = trait_num, colour = as.factor(trait_num), label = trait_num), show.legend = FALSE)
  pp2 %>% plotly::ggplotly(tooltip = "label")
```


# Plot coverage vs (mean) heritability 

```{r, eval = FALSE}
pp2 <- pp %>%
  dplyr::group_by(trait_num) %>%
  dplyr::summarise(mean_herit = mean(mean_herit), coverage = mean(in_interval)) %>%
  ggplot2::ggplot() + ggplot2::geom_point(ggplot2::aes(x = mean_herit, y = coverage, colour = as.factor(trait_num))) 
pp2 %>%
  plotly::ggplotly()

```

## Check correlation between (mean) heritability and coverage


```{r, eval = FALSE}
summ <- pp %>%
  dplyr::group_by(trait_num) %>%
  dplyr::summarise(mean_herit = mean(mean_herit), coverage = mean(in_interval))
cor(summ$mean_herit, summ$coverage)
```


# References





