---
title: "Cross-validation-plus for UKB traits"
author: "Frederick J. Boehm"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    number_sections: false
bibliography: ../Rmd/my.bib
params: 
  traits: 1
---

```{r}
library(magrittr)
```


## Custom R code

It should be straightforward to use DBSLMM outputs and fam files to write custom code for CV+.

Inputs needed: 

- fam file to get true values for $Y$. We need trait values for both the verification and training sets.    
- DBSLMM-based predictions for a trait, for all subjects and for each of the five folds that Sheng created. 




```{r, echo = TRUE, message = FALSE}
#(trait_type <- params$trait_type)
traits <- params$traits
trait_type <- "binary"
nfold <- 5
alpha <- 0.1
```






```{r}
(fbstr <- paste0("/net/mulan/disk2/fredboe/research/ukb-intervals/study_nfolds/", nfold, "-fold/"))
# read fam file to get ids vector
fam_file <- paste0(fbstr, "plink_file/ukb/binary/chr1.fam")
fam <- vroom::vroom(file = fam_file, col_names = FALSE)
#####
# make list output object to be filled
out <- list()
#if (trait_type == "continuous") {
#  	res_file <- paste0(fbstr, "res-continuous.rds")
#} else {
  	res_file <- paste0(fbstr, "res-binary", traits, ".rds")
#}
if (!file.exists(res_file)) {
  	res <- lapply(traits, function(x) {
		knitr::knit_child(
		'cv-plus-child.Rmd', 
		envir = environment(), 
		quiet = TRUE
		)
    get("out")
	})
  	saveRDS(res, file = res_file)
} else {
  	res <- readRDS(res_file)
}
```

```{r, eval = FALSE}
coverages <- lapply(X = res, 
                    FUN = function(mytib){
  mytib %>%
	dplyr::summarise(coverage = mean(in_interval))
       }
)
cov1k <- lapply(X = res, 
                    FUN = function(mytib){
  mytib %>%
    dplyr::summarise(coverage = mean(in_interval[1:1000]))
       }
) 
cov10k <- lapply(X = res, 
                    FUN = function(mytib){
  mytib %>%
    dplyr::summarise(coverage = mean(in_interval[1:10000]))
       }
) 

coverage_tib <- do.call("rbind", coverages) %>%
  tibble::as_tibble() %>%
  dplyr::mutate(trait_num = seq_along(res))
coverage_tib %>%
  gt::gt()

```

```{r, eval = FALSE}
cov_tib <- tibble::tibble(verif_size = c(length(res[[1]]$in_interval), 10000, 1000), 
                         coverage = c(unlist(coverages), unlist(cov10k), unlist(cov1k)))
cov_tib %>%
  gt::gt()
```

## Boxplots of interval width

```{r}
for (i in seq_along(res)){
  res[[i]] <- res[[i]] %>%
    dplyr::mutate(trait_num = i)
}
dplyr::bind_rows(res) %>%
  ggplot2::ggplot() + 
  ggplot2::geom_boxplot(ggplot2::aes(x = trait_num, y = interval_width, group = trait_num))
```

## Incorporate estimated heritability of traits

```{r, eval = FALSE}
if (trait_type == "continuous"){
  h_tib <- readr::read_csv("/net/mulan/disk2/fredboe/research/ukb-intervals/shell_scripts/h2.csv", 
                          col_names = FALSE) %>% 
            dplyr::select(-26) # drop 26, which is due to presence of trailing commas
} else {
  h_tib <- readr::read_csv("/net/mulan/disk2/research/ukb-intervals/shell_scripts/h2-binary.csv", 
                          col_names = FALSE) %>%
            dplyr::select(- tidyselect::last_col()) # drop 26, which is due to presence of trailing commas
}

pp <- h_tib %>% 
  dplyr::summarise(dplyr::across(.fns = mean)) %>%
  tidyr::pivot_longer(seq_along(traits)) %>%
  dplyr::rename(mean_herit = value) %>%
  dplyr::mutate(trait_num = traits) %>%
  dplyr::select(-name) %>%
  dplyr::select(trait_num, mean_herit) %>%
  dplyr::left_join(dplyr::bind_rows(res), by = "trait_num") 
pp2 <- pp %>%
  ggplot2::ggplot() + ggplot2::geom_boxplot(ggplot2::aes(x = mean_herit, y = interval_width, group = trait_num, colour = as.factor(trait_num), label = trait_num), show.legend = FALSE)
  pp2 %>% plotly::ggplotly(tooltip = "label")
```


# Plot coverage vs (mean) heritability 

```{r, eval = FALSE}
pp2 <- pp %>%
  dplyr::group_by(trait_num) %>%
  dplyr::summarise(mean_herit = mean(mean_herit), coverage = mean(in_interval)) %>%
  ggplot2::ggplot() + ggplot2::geom_point(ggplot2::aes(x = mean_herit, y = coverage, colour = as.factor(trait_num))) 
pp2 %>%
  plotly::ggplotly()

```

## Check correlation between (mean) heritability and coverage


```{r, eval = FALSE}
summ <- pp %>%
  dplyr::group_by(trait_num) %>%
  dplyr::summarise(mean_herit = mean(mean_herit), coverage = mean(in_interval))
cor(summ$mean_herit, summ$coverage)
```


# References





